---
title: "Odds Ratio vs Proportional odds"
author: "Jamie Ayton"
date: "18 May 2017"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library('tidyverse')
library('knitr')

```


### Summary

I have previously shown how to calculate the [Odds Ratio](http://htmlpreview.github.io/?https://github.com/jamieayton/odds-ratio/blob/master/odds-ratio.html) and how to estimate the true odds (and thus the true probabilities) from given Bookmakers odds.
Lets compare the 'Odds Ratio' method to another method of estimating the true odds - 'Proportional Odds.  

We will compare the two methods by calculating the implied true probabilities for each method and then calculating Ranked Probability Score for real odds for the 1x2 market of a selection of football matches. The odds data is from [www.football-data.co.uk](http://www.football-data.co.uk/).  

***


### Definitions

First lets re-cap how we define the 'Odds Ratio' and define 'Proportional Odds'.

For a three-way market, given bookmakers odds `1/p, 1/q, 1/r` with corresponding probabilities `p, q, r`; we wish to find true odds `1/x, 1/y, 1/z` with corresponding probabilities `x, y, z` such that $x + y + z = 1$.  

* Odds Ratio:  
    We defined the Odds Ratio of `x` & `p` to be $OR(x,p)$ and we find `x, y, z` such that $OR(x,p) = OR(y,q) = OR(z,r)$.

* Proportional Odds:  
    Let $v = p + q + r$, then we define $(x, y, z) := (\ p/v,\ q/v,\ r/v \ )$. 


***


### Data

```{r import data, echo=FALSE}

# read csv
football_data <- read_csv("C:/R Projects/odds-ratio/football_data.csv")



# filter for leagues & seasons
# select columns which are needed
# make sure odds columns are numeric & check that odds values are real ie. > 1
# remove NA rows

football_data <- football_data %>% 
  filter(
    season_code %in% c("09-10", "10-11", "11-12", "12-13", "13-14", "14-15", "15-16", "16-17")
  ) %>% 
  filter(
    league_code %in% c("D1", "E0", "E1", "E2", "E3", "F1", "I1", "SC0", "SP1")
  ) %>% 
  select(
    Div, Date, HomeTeam, AwayTeam, FTR, BbMxH, BbMxD, BbMxA, BbAvH, BbAvD, BbAvA, B365H, B365D, B365A, PSH, PSD, PSA
  ) %>% 
  mutate_at(
    .cols = vars(BbMxH, BbMxD, BbMxA, BbAvH, BbAvD, BbAvA, B365H, B365D, B365A, PSH, PSD, PSA), 
    .funs = funs(ifelse(as.numeric(.) >= 1L, as.numeric(.), NA))
  ) %>% 
  na.omit()
  



# bookmakers to use
bookmaker_codes <- tibble(
  bookmaker_name = c("Bet Brain - Max", "Bet Brain - Average", "Bet365", "Pinnacle Sports"), 
  bookmaker_code = c("BbMx", "BbAv", "B365", "PS")
)



# gather odds together to single column, 
# split out bookmaker & odds selection
# spread to seperate columns for H, D, A
# add column for 'overround'
# merge bookmaker_codes
# gather columns H, D, A again as 'odds_type'


gather_columns <- c("BbMxH", "BbMxD", "BbMxA", "BbAvH", "BbAvD", "BbAvA", "B365H", "B365D", "B365A", "PSH", "PSD", "PSA")
gather_columns <- which(colnames(football_data) %in%  gather_columns)

football_data <- football_data %>% 
  gather(
    ., "bookmaker_code","odds_value", gather_columns
  ) %>% 
  mutate(
    odds_type = substring(bookmaker_code, nchar(bookmaker_code)), 
    bookmaker_code = substring(bookmaker_code, 1, nchar(bookmaker_code)-1)
  ) %>% 
  mutate(
    odds_value = as.numeric(odds_value)
  ) %>% 
  spread(
    ., odds_type, odds_value
  ) %>% 
  mutate(
    overround = 1/H + 1/D + 1/A
  ) %>% 
  left_join(
    ., bookmaker_codes, by="bookmaker_code"
  ) %>% 
  gather(
    ., odds_type, original_odds, which(colnames(.) %in% c("H", "D", "A"))
  )


# rm no longer required items
rm(bookmaker_codes, gather_columns)


```



```{r add odds ratio, echo=FALSE}

# odds ratio solver function

three_way_market_solver <- function(odds_1, odds_2, odds_3, overround){
  
  # check inputs are all numeric & length == 1
  stopifnot(all(is.numeric(odds_1), is.numeric(odds_2), is.numeric(odds_3), is.numeric(overround)))
  stopifnot(all(length(odds_1) == 1, length(odds_2) == 1, length(odds_3) == 1, length(overround) == 1))
  
  # adjust odds to probabilities
  p <- 1/odds_1
  q <- 1/odds_2
  r <- 1/odds_3
  
  # solver function: minimise 'x + y - v' to find c such that x + y = v
  find_c <- function(c){
    abs((c*p/(1-p+c*p)) + (c*q/(1-q+c*q)) + (c*r/(1-r+c*r)) - overround)
  }
  
  # use optim to solve with lower bound to prevent -ve roots
  optimised_solver <- optim(
    par = c(1), fn = find_c, method = "Brent", lower = 1e-03, upper = 1e+03
  )
  
  # output parameter from optim
  optimised_solver$par[1]
}



# spread odds to give seperate columns for H,D,A
# apply three_way_market_solver to find value 'c' such that overround = 1
# use 'c' to define odds_ratio 'true' odds
# rm no longer needed columns

football_data <- football_data %>% 
  spread(
    ., odds_type, original_odds
  ) %>% 
  mutate(
    c = pmap_dbl(list(.$H, .$D, .$A, rep(1, nrow(.))), three_way_market_solver)
  ) %>% 
  gather(
    ., odds_type, original_odds, which(colnames(.) %in% c("H", "D", "A"))
  ) %>% 
  mutate(
    original_prob = 1/original_odds, 
    ratio_prob = c*original_prob/(1-original_prob+c*original_prob), 
    ratio_odds = 1/ratio_prob
  ) %>% 
  select(
    -one_of(c("c","original_prob", "ratio_prob"))
  )



```



```{r add proportional odds, echo=FALSE}

# add proportional odds
football_data <- football_data %>% 
  mutate(
    proportional_odds = original_odds * overround
  )



```

