---
title: "Odds Ratio"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library('tidyverse')
library('knitr')

```

### Summary

A simple recreation of the table shown in Keith Cheung's [Fixed-Odds Betting and Traditional Odds]("http://www.sportstradingnetwork.com/article/fixed-odds-betting-traditional-odds/") article.  
This includes a function to find the value 'c' when the odds ratios are equal for a two-way & three-way market.  

***

### Definitions

For a two-way betting market we will use the following definitions:  

* `1/p` & `1/q` are 'true' odds; where `p` & `q` are the 'true' probabilities of events `p` & `q` occurring
* similarly `1/x` & `1/y` are bookmakers odds; where `x` & `y` are the implied probabilities
* `v` is the overround of the bookmakers market and thus $v \ge 1$


Note that:

* For a three-way market we will use terms `p, q & r` for 'true' probabilities & `x, y & z` for the bookmakers probabilities
* We have $p + q = 1$ (or $p + q + r = 1$ for a three-way market)
* We have $x + y = v$ (or $x + y + z = v$)

***

### Two-Way Market Table


```{r two-way market table, echo=FALSE}


# create table
two_way_market_table <- tibble(
  `1/p` = c(seq(from = 2, to = 1.08, by = -0.04), 1.06, 1.04, 1.02, 1.01)  # by -0.04, for > 1.08; more granular < 1.08
  ) %>% 
  mutate(
    `1/q` = 1/(1-(1/`1/p`)), 
    v = 1.081  # v = overround
  ) %>% 
  mutate(
    p = 1/`1/p`,          # add p & q for easier calculations later
    q = 1/`1/q` 
  )


# function to find c
two_way_market_Solver <- function(odds_p, odds_q, overround){
  
  # check inputs are all numeric
  stopifnot(is.numeric(odds_p) && is.numeric(odds_q) && is.numeric(overround))
  
  # adjust odds to probabilities
  p <- 1/odds_p
  q <- 1/odds_q
  
  # solver function: x + y - v = 0 
  solver_function <- function(c){
    abs((c*p/(1-p+c*p)) + (c*q/(1-q+c*q)) - overround)
  }
  
  # use optim to solve with lower bound to prevent -ve roots
  optimised_solver <- optim(par = c(1), fn = solver_function, method = "Brent", lower = 1e-03, upper = 1e+03)
  
  # output parameter from optim
  optimised_solver$par[1]
  
}

# derive c with solver, add xy, yields & odds ratios
two_way_market_table <- two_way_market_table %>% 
  mutate(
    c = pmap_dbl(list(two_way_market_table$`1/p`, two_way_market_table$`1/q`, two_way_market_table$v), two_way_market_Solver)
  ) %>% 
  mutate(
    x = c*p/(1-p+c*p),
    y = c*q/(1-q+c*q), 
    `1/x` = 1/x, 
    `1/y` = 1/y
  ) %>% 
  mutate(
    yield_x = p*`1/x` - 1, 
    yield_y = q*`1/y` - 1, 
    odds_ratio_x_p = (x/(1-x)) / (p/(1-p)), 
    odds_ratio_y_q = (y/(1-y)) / (q/(1-q)) 
  )


# collect columns in the order as shown on the webpage
two_way_market_table <- two_way_market_table %>% 
  select(`1/p`, `1/q`, `1/x`, `1/y`, v, yield_x, yield_y, odds_ratio_x_p, odds_ratio_y_q)

# table output
kable(
  two_way_market_table, 
  col.names = c("1/p", "1/q", "1/x", "1/y", "v", "Y(x)", "Y(y)", "OR(x,p)", "OR(y,q)"), 
  digits = c( 3, 3, 3, 3, 4, 3, 3, 3, 3)
)


```

