---
title: "True Odds Estimation"
subtitle: "A Comparison between 'Proportional Odds', the 'Odds Ratio' and the 'Power Method'"
author: "Jamie Ayton"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library('tidyverse')
library('stringr')
library('ggplot2')
library('knitr')
library('grid')
library('gridExtra')  # for side-by-side plots

```


### Summary

I have previously shown how to calculate the [Odds Ratio](http://htmlpreview.github.io/?https://github.com/jamieayton/odds-ratio/blob/master/odds-ratio.html) and how to estimate the true odds (and thus the true probabilities) from given Bookmakers odds.
Lets compare the 'Odds Ratio' method to a couple of other methods of estimating the true odds - 'Proportional Odds' & the 'Power Method'.  

We will compare the three methods by calculating the implied true probabilities for each method and then calculating Ranked Probability Score for real bookmakers odds for the 1x2 market of a selection of football matches. The odds data is from [www.football-data.co.uk](http://www.football-data.co.uk/).  

***


### Definitions

First lets fully define the different methods.  

For a three-way market, given bookmakers odds `1/p, 1/q, 1/r` with corresponding probabilities `p, q, r`; we wish to find true odds `1/x, 1/y, 1/z` with corresponding probabilities `x, y, z` such that $x + y + z = 1$.  

* Proportional Odds:  
    Let $v = p + q + r$, then we define $(x, y, z) := (\ p/v,\ q/v,\ r/v \ )$.

* Odds Ratio:  
    We defined the Odds Ratio of `x` & `p` to be $OR(x,p)$ and we find `x, y, z` such that $OR(x,p) = OR(y,q) = OR(z,r)$.

* Power Odds:  
    Define $(x, y, z) := (\ p^{-\lambda},\ q^{-\lambda},\ r^{-\lambda} \ )$, where $\lambda \in \mathbb{R}$ is chosen such that $x + y + z = 1$.


***


### Data

```{r import data, include=FALSE}

# read csv
football_data <- read_csv("C:/R Projects/odds-ratio/football_data.csv")



# filter for leagues & seasons
# select columns which are needed
# make sure odds columns are numeric & check that odds values are real ie. > 1
# remove NA rows

football_data <- football_data %>% 
  filter(
    season_code %in% c("09-10", "10-11", "11-12", "12-13", "13-14", "14-15", "15-16", "16-17")
  ) %>% 
  filter(
    league_code %in% c("D1", "E0", "E1", "E2", "E3", "F1", "I1", "SC0", "SP1")
  ) %>% 
  select(
    league_code, season_code, Date, HomeTeam, AwayTeam, FTR, BbMxH, BbMxD, BbMxA, BbAvH, BbAvD, BbAvA, B365H, B365D, B365A, PSH, PSD, PSA
  ) %>% 
  mutate_at(
    .cols = vars(BbMxH, BbMxD, BbMxA, BbAvH, BbAvD, BbAvA, B365H, B365D, B365A, PSH, PSD, PSA), 
    .funs = funs(ifelse(as.numeric(.) >= 1L, as.numeric(.), NA))
  ) %>% 
  drop_na() %>% 
  mutate(
    match_id = seq(from=1, to=nrow(.), by=1)
  )


# bookmakers to use
bookmakers <- tibble(
  bookmaker_name = c("Bet Brain - Max", "Bet Brain - Average", "Bet365", "Pinnacle Sports"), 
  bookmaker_code = c("BbMx", "BbAv", "B365", "PS")
)

# leagues
leagues <- tibble(
  league_code = c("D1", "E0", "E1", "E2", "E3", "F1", "I1", "SC0", "SP1"), 
  league_name = c("German Bundesliga", "English Premier League", "English Championship", 
                  "English League One", "English League Two", "French Ligue One", 
                  "Italian Serie A", "Scottish Premiership", "Spanish La Liga" 
  )
)



# gather odds together to single column, 
# split out bookmaker & odds selection
# spread to seperate columns for H, D, A
# add column for 'overround'
# merge bookmaker_codes
# merge league_codes
# gather columns H, D, A again as 'odds_type'

gather_columns <- c("BbMxH", "BbMxD", "BbMxA", "BbAvH", "BbAvD", "BbAvA", "B365H", "B365D", "B365A", "PSH", "PSD", "PSA")
gather_columns <- which(colnames(football_data) %in%  gather_columns)

football_data <- football_data %>% 
  gather(
    ., "bookmaker_code","odds_value", gather_columns
  ) %>% 
  mutate(
    odds_type = substring(bookmaker_code, nchar(bookmaker_code)), 
    bookmaker_code = substring(bookmaker_code, 1, nchar(bookmaker_code)-1)
  ) %>% 
  mutate(
    odds_value = as.numeric(odds_value)
  ) %>% 
  spread(
    ., odds_type, odds_value
  ) %>% 
  mutate(
    overround = 1/H + 1/D + 1/A
  ) %>% 
  left_join(
    ., bookmakers, by="bookmaker_code"
  ) %>% 
  left_join(
    ., leagues, by="league_code"
  ) %>% 
  gather(
    ., odds_type, original_odds, which(colnames(.) %in% c("H", "D", "A"))
  )


# rm no longer required items
rm(bookmakers, leagues, gather_columns)

```



```{r number of observations, echo=FALSE}

# number of observations
n_matches <- football_data %>% select(Date, HomeTeam, AwayTeam) %>% unique(.) %>% summarise(count=n()) %>% as.integer()
n_bookmakers <- football_data %>% select(bookmaker_name) %>% unique(.) %>% summarise(count=n()) %>% as.integer()
n_odds <- n_matches*n_bookmakers


```


As mentioned the data is from [www.football-data.co.uk](http://www.football-data.co.uk/) and I have applied the following filters.  

* The leagues selected are: *`r str_c(unique(football_data$league_name), sep="", collapse=", ")`*  
* The seasons selected are: *`r str_c(paste0("20", unique(football_data$season_code)), sep="", collapse=", ")`*  
* The odds are from the following bookmakers: *`r str_c(unique(football_data$bookmaker_name), sep="", collapse=", ")`*  
* If any of the odds are missing then all the data for that specific game is removed from the dataset.
* If any of the odds values are invalid (ie. odds that are less than 1), again, all the data for that specific game is removed.


That leaves us with `r format(n_matches, nsmall=1, big.mark=",")` matches across `r format(n_bookmakers, nsmall=1, big.mark=",")` 
different bookmakers; giving us `r format(n_odds, nsmall=1, big.mark=",")` individual odds values.  


***

### RPS Comparison


```{r add proportional odds, echo=FALSE}

# add proportional odds
proportional_odds <- football_data %>% 
  select(
    match_id, bookmaker_code, odds_type, original_odds, overround
  ) %>% 
  mutate(
    proportional_odds = original_odds * overround
  )

```


```{r add odds ratio, echo=FALSE}

# odds ratio solver function

odds_ratio_solver <- function(odds_1, odds_2, odds_3, overround){
  
  # check inputs are all numeric, are finite & length == 1
  stopifnot(all(length(odds_1) == 1, length(odds_2) == 1, length(odds_3) == 1, length(overround) == 1))
  stopifnot(all(is.finite(odds_1), is.finite(odds_2), is.finite(odds_3), is.finite(overround)))
  
  # adjust odds to probabilities
  p <- 1/odds_1
  q <- 1/odds_2
  r <- 1/odds_3
  
  # solver function: minimise 'x + y - v' to find c such that x + y = v
  find_c <- function(c){
    abs((c*p/(1-p+c*p)) + (c*q/(1-q+c*q)) + (c*r/(1-r+c*r)) - overround)
  }
  
  # use optim to solve with lower bound to prevent -ve roots
  optimised_solver <- optim(
    par = c(1), fn = find_c, method = "Brent", lower = 1e-03, upper = 1e+03
  )
  
  # output parameter from optim
  optimised_solver$par[1]
}



# spread odds to give seperate columns for H,D,A
# apply odds_ratio_solver to find value 'c' such that overround = 1
# use 'c' to define odds_ratio 'true' odds
# rm no longer needed columns

odds_ratio <- football_data %>% 
  select(
    match_id, bookmaker_code, odds_type, original_odds, overround
  ) %>% 
  spread(
    ., odds_type, original_odds
  ) %>% 
  mutate(
    c = pmap_dbl(list(.$H, .$D, .$A, rep(1, nrow(.))), odds_ratio_solver)
  ) %>% 
  gather(
    ., odds_type, original_odds, which(colnames(.) %in% c("H", "D", "A"))
  ) %>% 
  mutate(
    original_prob = 1/original_odds, 
    ratio_prob = c*original_prob/(1-original_prob+c*original_prob), 
    ratio_odds = 1/ratio_prob
  ) %>% 
  select(
    -one_of(c("c","original_prob", "ratio_prob"))
  )



```



```{r add power odds, echo=FALSE}


# solver function
power_odds_solver <- function(odds_1, odds_2, odds_3){
  
  # check inputs are all numeric & length == 1
  stopifnot(all(is.numeric(odds_1), is.numeric(odds_2), is.numeric(odds_3)))
  stopifnot(all(length(odds_1) == 1, length(odds_2) == 1, length(odds_3) == 1))
  
  # adjust odds to probabilities
  p <- 1/odds_1
  q <- 1/odds_2
  r <- 1/odds_3
  
  # find lambda function
  find_lambda <- function(lambda){
      abs(p^(lambda) + q^(lambda) + r^(lambda) - 1)
  }
  
  # parameter range
  min_par <- log(1/3)/log(min(p,q,r))
  max_par <- log(1/3)/log(max(p,q,r))
  
  # use optim to solve with lower bound to prevent -ve roots
  optimised_solver <- optimize(find_lambda, c(min_par, max_par), tol = 0.00001)
  
  # output parameter from optim
  optimised_solver$minimum[1]
}


# spread odds to give seperate columns for H,D,A
# apply power_odds_solver to find value 'lambda' such that overround = 1
# use 'lambda' to define power odds 'true' odds
# rm no longer needed columns

power_odds <- football_data %>% 
  select(
    match_id, bookmaker_code, odds_type, original_odds, overround
  ) %>% 
  spread(
    ., odds_type, original_odds
  ) %>% 
  mutate(
    lambda = pmap_dbl(list(.$H, .$D, .$A), power_odds_solver), 
    precision = abs((1/H)^lambda + (1/D)^lambda + (1/A)^lambda - 1)
  ) %>% 
  gather(
    ., odds_type, original_odds, which(colnames(.) %in% c("H", "D", "A"))
  ) %>% 
  mutate(
    original_prob = 1/original_odds, 
    power_prob = original_prob^(lambda), 
    power_odds = 1/power_prob
  ) %>% 
  select(
    -one_of(c("lambda", "precision", "original_prob", "power_prob"))
  )


```


```{r combine odds estimates, echo=FALSE}

# combine odds, then gather odds method & value
odds_comparison <- reduce(
    list(football_data, proportional_odds, odds_ratio, power_odds), 
    left_join, 
    by=c("match_id", "bookmaker_code", "odds_type", "original_odds", "overround")
  ) %>% 
  gather(
    ., odds_method, odds_value, which(colnames(.) %in% c("original_odds", "proportional_odds", "ratio_odds", "power_odds"))
  )

rm(proportional_odds, odds_ratio,power_odds)


```


```{r rps calculation, echo=FALSE}

# spread odds_type so that H, D & A odds are in individual cols
# calculate RPS

rps_comparison <- odds_comparison %>% 
  spread(., odds_type, odds_value) %>% 
  mutate(rps = 0.5*((ifelse(FTR=="H", 1, 0) - 1/H)^2 + (ifelse(FTR=="D", 1, 0) - 1/D)^2 + (ifelse(FTR=="A", 1, 0) - 1/A)^2))



```





Below is a table comparing the calculated RPS values (lower is better) across the differents true odds estimates for the  various different bookmakers.

```{r bookmakers comparison, echo=FALSE}

# group by odds_method & bookmakers
# summarize rps by groupings (incl count to ensure constant across the board)
# spread to give odds methods as individual cols of rps values
# re-order cols

# bookmakers comparison - kable output
bookmakers_summary <- rps_comparison %>% 
  bind_rows(., rps_comparison %>% mutate(bookmaker_name = c("Total"))) %>% 
  group_by(odds_method, bookmaker_name) %>% 
  summarise(
    count = n(), 
    overround = mean(overround), 
    rps = mean(rps)
  ) %>% 
  spread(
    odds_method, rps
  ) %>% 
  select(bookmaker_name, count, overround, original_odds, proportional_odds, ratio_odds, power_odds)



# table output
kable(
  bookmakers_summary, 
  col.names = c("Bookmaker", "Matches", "Overround", "Original Odds", "Proportional Odds", "Ratio Odds", "Power Odds"), 
  digits = c(0,6,3,6,6,6,6), 
  align = 'l', 
  format = "markdown", 
  caption = "RPS Summary by Bookmaker & Odds Estimation Method"
)


```

We see that in general the RPS value is lowest for the 'Power Odds', then 'Ratio Odds', 'Proportional Odds' and finally the 'Original Odds' have the highest RPS values. This means that in general that the 'Power' method produces odds which are closest to the True Odds of all the methods.  
Curiously this is not true for the bookmaker 'Bet Brain - Max', perhaps this is due to the very low overround.  
We also note that the bookmaker 'Pinnacle Sports' consistently has the lowest RPS regardless of the odds method.  


***



### Pinnacle Sports Odds

Let's focus on the 'Pinnacle Sports' odds and try to look at the different odds methods in more detail. 

&nbsp; 
  
### Odds Comparison

First lets visually compare the odds to the original odds.  


```{r plot - odds & probs comparison, echo = FALSE, message = FALSE, fig.width = 10, fig.height = 5, fig.show='hold', fig.align='center'}

# plot original odds vs new odds for each method
plot_data <- odds_comparison %>% 
  filter(bookmaker_code == "PS") %>% 
  select(match_id, odds_type, odds_method, odds_value) %>% 
  spread(odds_method, odds_value) %>% 
  gather(odds_method, odds_value, which(colnames(.) %in% c("proportional_odds", "ratio_odds", "power_odds")))


# loess fit for probs
loess_probs_comparison <- plot_data %>% 
  mutate(original_probs = 1/original_odds, probs_value = 1/odds_value) %>% 
  select(original_probs, probs_value, odds_method) %>% 
  group_by(odds_method) %>% 
  do(fit = loess(probs_value ~ original_probs, .))


# data for plots
plot_odds <- tibble(original_probs = seq(0, 1, by=0.00001)) %>% 
  mutate(
    original_odds = 1/original_probs
  ) %>% 
  mutate(
    ratio_odds = as.vector(predict(loess_probs_comparison$fit[loess_probs_comparison$odds_method == "ratio_odds"][[1]], newdata = .)), 
    power_odds = as.vector(predict(loess_probs_comparison$fit[loess_probs_comparison$odds_method == "power_odds"][[1]], newdata = .)), 
    proportional_odds = as.vector(predict(loess_probs_comparison$fit[loess_probs_comparison$odds_method == "proportional_odds"][[1]], newdata = .))
  ) %>% 
  drop_na() %>% 
  mutate(
    ratio_odds = 1/ratio_odds, 
    power_odds = 1/power_odds, 
    proportional_odds = 1/proportional_odds
  ) %>% 
  gather(odds_method, odds_value, which(colnames(.) %in% c("ratio_odds", "power_odds", "proportional_odds"))) %>% 
  group_by(odds_method)

plot_probs <- tibble(original_probs = seq(0, 1, by=0.00001)) %>% 
  mutate(
    ratio_odds = as.vector(predict(loess_probs_comparison$fit[loess_probs_comparison$odds_method == "ratio_odds"][[1]], newdata = .)), 
    power_odds = as.vector(predict(loess_probs_comparison$fit[loess_probs_comparison$odds_method == "power_odds"][[1]], newdata = .)), 
    proportional_odds = as.vector(predict(loess_probs_comparison$fit[loess_probs_comparison$odds_method == "proportional_odds"][[1]], newdata = .))
  ) %>% 
  drop_na() %>% 
  gather(odds_method, probs_value, which(colnames(.) %in% c("ratio_odds", "power_odds", "proportional_odds"))) %>% 
  group_by(odds_method)


# smooth line plot - shows variation at tail for large odds values, but doesn't show differences at small odds values
plot_odds <- plot_odds %>% 
  ggplot(aes(x = original_odds, y = odds_value, color = odds_method)) + 
  geom_line() + 
  geom_abline(slope = 1, intercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(1, 50), ylim = c(1, 50)) + 
  scale_colour_discrete(
    name = "Odds Method",
    breaks = c("power_odds", "proportional_odds", "ratio_odds"),
    labels = c("Power Odds", "Proportional Odds", "Ratio Odds")
  ) + 
  theme_classic() + 
  theme(legend.justification=c(1,0), legend.position=c(1,0)) + 
  labs(x = "Original Odds \n Fig 1.1", y = "Estimated True Odds") + 
  theme(plot.title = element_text(hjust = 0.5))

# smooth line plot - probability scale
plot_probs <- plot_probs %>% 
  ggplot(aes(x = original_probs, y = probs_value, color = odds_method)) + 
  geom_line() + 
  geom_abline(slope = 1, intercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) + 
  scale_colour_discrete(
    name = "Odds Method",
    breaks = c("power_odds", "proportional_odds", "ratio_odds"),
    labels = c("Power Odds", "Proportional Odds", "Ratio Odds")
  ) + 
    theme_classic() + 
  theme(legend.justification=c(1,0), legend.position=c(1,0)) + 
  labs(x = "Original Probability \n Fig 1.2", y = "Estimated True Probability")


grid.arrange(plot_odds, nullGrob(), plot_probs,  top = textGrob("A Comparison of the Different Odds Methods"), widths=c(0.475, 0.05, 0.475))

# rm
rm(loess_probs_comparison)
rm(plot_odds, plot_probs)


```


* Fig 1.1 : 
    + For large odds values the differences between the different methods & the $y = x$ dotted line becomes more apparent. We see that the 'Estimated Odds' are larger however the 'Ratio Odds' & 'Power Odds' diverge away above the dotted line.
    + For small odds values it is very hard to distinguish the methods from the 'Original Odds' and from each other. 
    + The majority of the data falls within the 'small odds values' range - it is clear that we are not going to learn anything from plotting 'Odds vs Odds'.
* Fig 1.2 : 
    + Converting to a 'Probability vs Probability' plot is actually worse than the 'Odds vs Odds' plot! This time we can't see if there are any differences at all. We will need to come up with another way to plot the data.  
  

&nbsp;


Since Fig 1.1 shows a general increase for large odds value - lets try plotting the change in the value of the odds in the form of a scalar value, where the scalar multiple $\lambda$ is such that $\text{Estimated True Odds} = \lambda * \text{Original Odds}$.  

&nbsp;

```{r plot - odds & probs scalar increase , echo = FALSE, message = FALSE, fig.width = 10, fig.height = 5, fig.show='hold', fig.align='center'}

# add scalar multiples
plot_data <- plot_data %>% 
  mutate(odds_scalar_multiple = odds_value/original_odds) %>% 
  mutate(probs_scalar_multiple = (1/odds_value)/(1/original_odds))


# loess fit for probs_scalar & odds_scalar
loess_probs_scalar <- plot_data %>% 
  mutate(original_probs = 1/original_odds) %>% 
  select(original_probs, probs_scalar_multiple, odds_method) %>% 
  group_by(odds_method) %>% 
  do(fit = loess(probs_scalar_multiple ~ original_probs, .))


# data for plots
plot_probs <- tibble(original_probs = seq(0, 1, by=0.00001)) %>% 
  mutate(
    original_odds = 1/original_probs
  ) %>% 
  mutate(
    ratio_odds = as.vector(predict(loess_probs_scalar$fit[loess_probs_scalar$odds_method == "ratio_odds"][[1]], newdata = .)), 
    power_odds = as.vector(predict(loess_probs_scalar$fit[loess_probs_scalar$odds_method == "power_odds"][[1]], newdata = .)), 
    proportional_odds = as.vector(predict(loess_probs_scalar$fit[loess_probs_scalar$odds_method == "proportional_odds"][[1]], newdata = .))
  ) %>% 
  drop_na() %>% 
  gather(odds_method, probs_scalar_multiple, which(colnames(.) %in% c("ratio_odds", "power_odds", "proportional_odds"))) %>% 
  group_by(odds_method)

plot_odds <- plot_probs %>% 
  mutate(odds_scalar_multiple = 1/probs_scalar_multiple) %>% 
  group_by(odds_method)


# plots - odds & probs 
plot_odds <- plot_odds %>% 
  ggplot(aes(original_odds, odds_scalar_multiple, color = odds_method)) + 
  geom_smooth() + 
  geom_hline(yintercept = 1, lty = 2) + 
  coord_cartesian(xlim = c(1, 10), ylim = c(1, 1.1)) + 
  scale_colour_discrete(
    name = "Odds Method",
    breaks = c("power_odds", "proportional_odds", "ratio_odds"),
    labels = c("Power Odds", "Proportional Odds", "Ratio Odds")
  ) + 
  theme_classic() + 
  theme(legend.justification = c(0.2, 0.8), legend.position = c(0.2, 0.8)) + 
  labs(x = "Original Odds \n Fig 1.3", y = "Scalar Multiple (Odds)")

plot_probs <- plot_probs %>% 
  ggplot(aes(original_probs, probs_scalar_multiple, color = odds_method)) + 
  geom_smooth() + 
  geom_hline(yintercept = 1, lty = 2) + 
  coord_cartesian(xlim = c(1, 0), ylim = c(0.75, 1)) + 
  scale_colour_discrete(
    name = "Odds Method",
    breaks = c("power_odds", "proportional_odds", "ratio_odds"),
    labels = c("Power Odds", "Proportional Odds", "Ratio Odds")
  ) + 
  theme_classic() + 
  theme(legend.justification = c(0.8, 0.2), legend.position = c(0.8, 0.2)) + 
  labs(x = "Original Probability \n Fig 1.4", y = "Scalar Multiple (Probability)")


# output plots side by side
grid.arrange(plot_odds, nullGrob(), plot_probs,  top = textGrob("A Comparison of the Different Odds Methods"), widths=c(0.475, 0.05, 0.475))

rm(loess_probs_scalar)
rm(plot_odds, plot_probs)

```

* Fig 1.3: Odds Plot
    + The fitted curves for the different methods are above the $y = 1$ dotted line. So the True Odds are greater than the Original Odds across the whole range.
    + We see that the scalar increase for the 'Proportional Odds' method is quite uniform across the whole range. So this method increases the 'Original Odds' by the same scalar amount across the whole odds range.
    + The Power Odds & Ratio Odds have a linear trend between the Original Odds value and the scalar multiple - so as the odds value increases the scalar multiple also increases. 
    + We have an intersection point at Original Odds of ~ 3.0.  
      Above this point the 'Power Odds' method produces the largest True Odds values and the 'Proportional Odds' method produces the smallest True Odds values.  
      Below this point the 'Power Odds' method produces the smallest True Odds values and the 'Proportional Odds' produces the largest True Odds values.  

* Fig 1.4: Probability Plot
    + This plot just shows the same as the Odds plot, however since probabilities are the inverse of the odds values our whole plot is inversed, including the scalar multiples.  


```{r plot - odds scalar increase facet by odds_type, include = FALSE, fig.width = 10, fig.height = 5, fig.show='hold', fig.align='center'}

# odds plot ~ facetted by odds_type
plot_data %>% 
  ggplot(aes(original_odds, odds_scalar_multiple, color = odds_method)) + 
  geom_smooth() + 
  geom_hline(yintercept = 1, lty = 2) + 
  facet_wrap(~odds_type) + 
  coord_cartesian(xlim = c(0, 5), ylim = c(1, 1.1)) + 
  scale_colour_discrete(
    name = "Odds Method",
    breaks = c("power_odds", "proportional_odds", "ratio_odds"),
    labels = c("Power Odds", "Proportional Odds", "Ratio Odds")
  ) + 
  theme_classic() + 
  theme(legend.justification = c(0.1, 0.9), legend.position = c(0.1, 0.9)) + 
  labs(x = "Original Odds \n Fig 1.3", y = "Scalar Multiple")

# nothing 'new' here - same trend as when all odds_types were grouped together

```



&nbsp;


So now we understand how the various methods differ in producing 'True Odds' across the range of our 'Original Odds', but how does this link to the RPS values which showed that the 'Power Odds' method was the most accurate.  

Lets visualize the RPS values as a comparison to the Original Odds values.  

&nbsp;

&nbsp;

### RPS Comparison

If we try to plot the RPS values we will run into the same situation that we did with visualizing the odds values - the values will be too close to distinguish the different odds methods. So similar to the odds plots we will need to plot a variable which represents the change in the RPS value.  

Lets define the RPS Difference $\lambda$ to be such that $\lambda = \text{RPS}_{\text{True Odds}} - \text{RPS}_{\text{Original Odds}}$. Then if $\lambda < 0$ this means that the 'True Odds' produced are more accurate than the 'Original odds', and conversely if $\lambda > 0$ then the 'Original odds' are more accurate.  

  
```{r - rps comparison diff, echo = FALSE, message = FALSE, fig.width = 12, fig.height = 6, fig.show='hold', fig.align='center'}

# tibble with rps values
plot_data_rps <- rps_comparison %>% 
  select(match_id, bookmaker_code, overround, odds_method, rps, H, D, A) %>% 
  filter(bookmaker_code == "PS") %>% 
  gather(odds_type, odds_value, which(colnames(.) %in% c("H", "D", "A")))

# seperate out original_odds and original_rps
# right join odds values and rps values for the other odds methods
# add original_probs & rps difference
plot_data_rps <- plot_data_rps %>% 
  filter(odds_method == "original_odds") %>% 
  select(match_id, bookmaker_code, overround, odds_type, original_odds = odds_value, original_rps = rps) %>% 
  right_join(., plot_data_rps %>% filter(odds_method != "original_odds"), by=c("match_id", "bookmaker_code", "overround", "odds_type")) %>% 
  mutate(
    original_probs = 1/original_odds, 
    rps_diff = rps - original_rps
  )

# fit loess model for plotting
loess_rps_diff <- plot_data_rps %>% 
  select(original_probs, rps_diff, odds_method) %>% 
  group_by(odds_method) %>% 
  do(fit = loess(rps_diff ~ original_probs, .))

# data for plot
plot_rps_diff <- tibble(original_probs = seq(0, 1, by=0.00001)) %>% 
  mutate(
    ratio_odds = as.vector(predict(loess_rps_diff$fit[loess_rps_diff$odds_method == "ratio_odds"][[1]], newdata = .)), 
    power_odds = as.vector(predict(loess_rps_diff$fit[loess_rps_diff$odds_method == "power_odds"][[1]], newdata = .)), 
    proportional_odds = as.vector(predict(loess_rps_diff$fit[loess_rps_diff$odds_method == "proportional_odds"][[1]], newdata = .))
  ) %>% 
  drop_na() %>% 
  gather(odds_method, rps_diff, which(colnames(.) %in% c("ratio_odds", "power_odds", "proportional_odds"))) %>% 
  group_by(odds_method)

# plot - probs vs rps_diff
plot_rps_diff <- plot_rps_diff %>% 
  ggplot(aes(original_probs, rps_diff, colour = odds_method)) + 
  geom_smooth() + 
  geom_hline(yintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(0, 1), ylim = c(-0.00025, 0)) + 
  scale_colour_discrete(
    name = "Odds Method",
    breaks = c("power_odds", "proportional_odds", "ratio_odds"),
    labels = c("Power Odds", "Proportional Odds", "Ratio Odds")
  ) + 
  theme_classic() + 
  theme(legend.justification = c(1, 0.8), legend.position = c(1, 0.8)) + 
  labs(x = "Original Probabilities \n Fig 1.5", y = "RPS Difference", title = "Comparing the Accuracy of Various Methods for Determining True Odds")

plot_rps_diff


rm(loess_rps_diff)


```

Lets breakdown the graph by each Odds Method: 

* Power Odds  
    + More accurate than the 'Original Odds' for the whole probability domain - since the RPS Difference value is negative.
    + Most accurate method for (approximately) Original Probability < ~0.25 & 0.5 < Original Probability.
    + Least accurate method for 0.3 < Original Probability < 0.4.
* Proportional Odds  
    + More accurate than the 'Original Odds' for 0.2 < Original Probability < 0.65.
    + Most accurate method for 0.3 < Original Probability < 0.4.
    + Least accurate method for Original Probability < 0.3 & 0.4 < Original Probability
* Ratio Odds  
    + More accurate than the 'Original Odds' for 0.1 < Original Probability < 0.8.
    + It appears there are two very small regions around Original Probability = 0.3 & Original Probability = 0.4 where this method is the most accurate method.
    + This method doesn't appear to be the least accurate method for any probability domain.



&nbsp;

  

### A Combined Method

What determines these cut off points where the most accurate method switches between the various odds methods?  

If we could determine these cut off points we could create a more accurate method for determining the True Odds by picking the most appropriate method given the Original Odds values.  

Lets consider some other factors which might effect these cut off points:  

1. Odds Type - for example, does the odds value relate to 'Home', 'Draw' or 'Away' match outcome?
2. Favourite - does it affect the cutoff if the Home team is the favourite or if the Away team is the favourite?
3. Overround - as the overround increase does this effect the cut off points or promote a different method?
4. Bookmaker - do we see the same patterns for all bookmakers?

  
In order to investigate these factors we need a method for determining these cut off points. We will do this by fitting a [LOESS Model]("https://en.wikipedia.org/wiki/Local_regression") using the Original Probabilities as our predictor variable (since it is defined regularly on the domain $\left[0, 1\right]$), and the RPS values or change in RPS values as a response variable. We can then use these predictions to determine the values where there are crossovers.



```{r - loess rps fit, echo = FALSE, message = FALSE}

# fit loess for rps_diff ~ original_odds
loess_odds_methods <- plot_data_rps %>% 
  select(match_id, original_odds, original_rps, odds_method, rps_diff) %>% 
  mutate(original_probs = 1/original_odds) %>% 
  group_by(odds_method) %>%  
  do(fit = loess(rps_diff ~ original_probs, .))


```


As an example - lets take the RPS Comparison from above & calculate the intersection points between the various methods.


```{r - rps comparison crossover points, echo = FALSE, message = FALSE, fig.width = 12, fig.height = 6, fig.show='hold', fig.align='center'}

# tibble with even lattice of points and predictions
rps_comparison_cross_over_points <- tibble(original_probs = seq(0, 1, by=0.00001)) %>% 
  mutate(
    ratio_odds = as.vector(predict(loess_odds_methods$fit[loess_odds_methods$odds_method == "ratio_odds"][[1]], newdata = .)), 
    power_odds = as.vector(predict(loess_odds_methods$fit[loess_odds_methods$odds_method == "power_odds"][[1]], newdata = .)), 
    proportional_odds = as.vector(predict(loess_odds_methods$fit[loess_odds_methods$odds_method == "proportional_odds"][[1]], newdata = .))
  ) %>% 
  drop_na()


# original plot
plot_rps_diff <- rps_comparison_cross_over_points %>% 
  gather(odds_method, rps_diff, which(colnames(.) %in% c("ratio_odds", "power_odds", "proportional_odds"))) %>% 
  ggplot(aes(original_probs, rps_diff, colour = odds_method)) + 
  geom_smooth() + 
  geom_hline(yintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(0, 1), ylim = c(-0.00015, 0)) + 
  scale_colour_discrete(
    name = "Odds Method",
    breaks = c("power_odds", "proportional_odds", "ratio_odds"),
    labels = c("Power Odds", "Proportional Odds", "Ratio Odds")
  ) + 
  theme_classic() + 
  theme(legend.justification = c(1, 0.6), legend.position = c(1, 0.6)) + 
  labs(x = "Original Probabilities \n Fig 1.6", y = "RPS Difference")


# find crossover/intersection points - first find location of min, then output the value and the method
rps_comparison_cross_over_points <- rps_comparison_cross_over_points %>% 
  mutate(
    min_odds = pmin(ratio_odds, power_odds, proportional_odds), 
    min_method = rps_comparison_cross_over_points %>% 
      select(one_of(c("ratio_odds", "power_odds", "proportional_odds"))) %>% transpose() %>% map_chr(function(x) names(which.min(x)))
  ) %>% 
  mutate(
    crossover = ifelse(lead(min_method) != min_method | lag(min_method) != min_method | row_number() %in% c(1,nrow(.)), 1, 0), 
    event_label = ifelse(crossover == 1, paste0("x = ", original_probs), "")
  )

# plot of min_odds curve
plot_rps_min <- rps_comparison_cross_over_points %>% 
  ggplot(aes(original_probs, min_odds)) + 
  geom_line(colour = "black") + 
  geom_hline(yintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(0, 1), ylim = c(-0.00015, 0))

# add cross over points
plot_rps_min <- plot_rps_min + 
  geom_vline(aes(xintercept = original_probs), data = rps_comparison_cross_over_points %>% filter(crossover == 1), lty = 2) + 
  geom_text(
    aes(original_probs, 0, label = event_label), size = 4, angle = 90, vjust = +1.2, hjust = 3, 
    data = rps_comparison_cross_over_points %>% filter(crossover == 1) %>% filter(row_number() %in% c(5, 9))
  ) + 
  geom_text(
    aes(original_probs, 0, label = event_label), size = 4, angle = 90, vjust = -0.6, hjust = 3, 
    data = rps_comparison_cross_over_points %>% filter(crossover == 1) %>% filter(row_number() %in% c(2, 6))
  ) 

# split data in two to prevent grouping of disjoint sets when applying geom_area
geom_area_breakpoint <- rps_comparison_cross_over_points %>% 
  filter(min_method == "proportional_odds" & crossover == 1)
geom_area_breakpoint <- min(geom_area_breakpoint$original_probs)

# add geom area
plot_rps_min <- plot_rps_min + 
  geom_area(aes(fill = min_method), alpha=0.1, data = rps_comparison_cross_over_points %>% filter(original_probs <= geom_area_breakpoint)) + 
  geom_area(aes(fill = min_method), alpha=0.1, data = rps_comparison_cross_over_points %>% filter(original_probs >= geom_area_breakpoint)) + 
  scale_fill_discrete(
    name = "Odds Method",
    breaks = c("power_odds", "proportional_odds", "ratio_odds"),
    labels = c("Power Odds", "Proportional Odds", "Ratio Odds")
  )

# add meta data
plot_rps_min <- plot_rps_min + 
  theme_classic() + 
  theme(legend.justification = c(0.8, 0.8), legend.position = c(0.8, 0.8)) + 
  labs(x = "Original Probabilities \n Fig 1.7", y = "RPS Difference")


# output plots side by side
grid.arrange(plot_rps_diff, nullGrob(), plot_rps_min,  top = textGrob("Combined Odds Method"), widths=c(0.475, 0.05, 0.475))

rm(plot_rps_diff, plot_rps_min)

```


From Fig 1.7 we see that we can define a new Odds Method by choosing a different method for different domains of the Original Probabilties:  

```{r final model, echo = FALSE, results = 'asis'}

# output above model as kable table

model_spec <- tibble(
  original_probs_lower = c(
    "0", 
    "0.27401", 
    "0.29634", 
    "0.44363", 
    "0.45983"
  ), 
  original_probs_equality = c(
    " < x <= ", 
    " < x <= ", 
    " < x <= ", 
    " < x <= ", 
    " < x <= "
  ), 
  original_probs_upper = c(
    "0.27401", 
    "0.29634", 
    "0.44363", 
    "0.45983", 
    "1.0"
  ),
  odds_method = c(
    "Power Odds", "Ratio Odds", "Proportional Odds", "Ratio Odds", "Power Odds"
  )
)


# table output
kable(
  model_spec, 
  col.names = c("Lower Bound", "Original Probabilities", "Upper Bound", "Odds Method"), 
  align = c("r", "c", "l", "c"), 
  padding = 0, 
  format = "markdown"
)

    

```

&nbsp; 

Finally - here is a table comparing the RPS values for the Original Odds method, the 3 True Odds methods already seen, and the Combined Method for determining True Odds. We already know that the Power Odds method is the most accurate of the 3 methods - and the Combined Method is more accurate but only by a very small amount.  


```{r rps calculation with combined method, echo=FALSE}

# add combined method
rps_comparison_with_combined <- rps_comparison %>% 
  filter(bookmaker_code == "PS") %>% 
  select(match_id, bookmaker_code, overround, odds_method, A, D, H, rps) %>% 
  gather(odds_type, odds_value, which(colnames(.) %in% c("H", "D", "A")))

combined_method <- rps_comparison_with_combined %>% 
  select(match_id, bookmaker_code, overround, odds_method, odds_type, odds_value) %>% 
  filter(odds_method == "original_odds") %>% 
  mutate(probs_value = 1/odds_value) %>% 
  mutate(
    combined_method = case_when(
      0 < probs_value & probs_value <= 0.27401 ~ "power_odds", 
      0.27401 < probs_value & probs_value <= 0.29634 ~ "ratio_odds", 
      0.29634 < probs_value & probs_value <= 0.44363 ~ "proportional_odds", 
      0.44363 < probs_value & probs_value <= 0.45983 ~ "ratio_odds", 
      0.45983 < probs_value & probs_value <= 1.0 ~ "power_odds"
    )
  ) %>% 
  left_join(
    ., 
    rps_comparison_with_combined %>% select("match_id", "bookmaker_code", "overround", "odds_type", "odds_value", "odds_method", "rps"),  
    by = c("match_id", "bookmaker_code", "overround", "odds_type", "combined_method" = "odds_method")
  ) %>% 
  rename(original_odds_value = odds_value.x, odds_value = odds_value.y) %>% 
  mutate(odds_method = "combined_method") %>% 
  select(match_id, bookmaker_code, overround, odds_method, odds_type, odds_value, rps)

# add combine_method to existing tibble of methods
rps_comparison_with_combined <- bind_rows(
  rps_comparison_with_combined, combined_method
)


# output summary table
rps_comparison_with_combined_summary <- rps_comparison_with_combined %>% 
  group_by(odds_method) %>% 
  summarise(
    matches = n_distinct(match_id), 
    overround = mean(overround), 
    rps = mean(rps)
  ) %>% 
  ungroup() %>%
  spread(
    odds_method, rps
  )
  
# table output
kable(
  rps_comparison_with_combined_summary, 
  col.names = c("Matches", "Overround", "Combined Method", "Original Odds", "Power Odds", "Proportional Odds", "Ratio Odds"), 
  digits = c(0,6,6,6,6,6,6), 
  align = 'l', 
  format = "markdown", 
  caption = "RPS Summary by Bookmaker & Odds Estimation Method"
)



```

&nbsp; 

***

### Further Investigations

There are other factors that could affect which odds method is the most accurate for predicting the 'True Odds':

* Overround
* Bookmaker - this factor is heavily linked to the 'Overround' since each bookmaker will run a different Overround %
* Which team is the favourite - ie. either Home team or the Away team
* The match outcome that is being predicted - for example the 'Draw' odds value is usually a midpoint between the other two odds values.

The above factors are all heavily linked together so it is difficult to analyse each individually. For example the 'Bookmaker' & 'Overround' factors are heavily linked since each bookmaker will run a different Overround %, but then the overround could also depend on the league that the odds are from since liquidity might be lower in this league.

&nbsp; 

***

### Finale 

In summary we see that:

* The 'Pinnacle Sports' odds are the most accurate set of Original Odds
* The Power Odds method is the most accurate overall method for predicting the 'True Odds' from the Original Odds
    - The Power Odds method is the most accurate method for 'large' & 'small' values of the Original Probabilities
    - The Proportional Odds method is the most accurate method for 'medium' values of the Original Probabilities
* We can create a combined method using the most accurate method for different domains of Original Probabilities
* There are lots of other factors that could effect which Odds method is the most accurate for predicting 'True Odds'

&nbsp; 

***

